\name{}: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-04-16 paper/main.tex
Add suplement tables and figures, captions still open, Added TODO for citation

\documentclass{bioinfo}
\usepackage[english]{babel}
\copyrightyear{2015} \pubyear{2015}
\usepackage{natbib}
\bibliographystyle{natbib}
\usepackage{url}

@@

Both approaches may be used in conjunction with network inference
\cite{}. Biological networks used for DTI prediction include
protein--protein interaction networks \cite{} and networks including
several other types of biological relations, including similarity
between represented entities \cite{Gottlieb2011}. Network-based DTI
prediction methods use the guilt-by-association principle
\cite{Oliver2000}
and assume that a protein is a likely target for a drug if many of the
protein's neighbors in the interaction network are targets of the drug
\cite{Gillis2012}. Network-based methods have been applied successfully
to DTI prediction. However, if DTIs are taken as direct physical
interactions between a drug and protein, it remains an unresolved
question whether the network-based guilt-by-association hypothesis is
true, or whether an interaction of a drug and protein dys-regulates
several of the protein's interaction partners, and therefore resulting
in downstream effects that are not direct interactions but only
downstream consequences of an interaction.

@@

As we want to learn from the similarity of drug side effects and
protein phenotypes, we use a deep half-twin neural network with a
contrastive loss using cosine similarity.  A half-twin neural network
aims to learn a similarity between two embeddings of the same
dimension.  The precomputed embeddings may have different
dimensionality and we first process them using a fully connected
neural network layer which takes as input an embedding and outputs a
representation of a particular size, i.e., we use this layer as a
trainable feature transformation and apply it to reduce the
representation size of the embeddings for drugs and proteins
separately.  An example structure for both types of features can be
found in Figure~\ref{fig:HalfTwinNetwork}.
% While a regular deep neural network, denoted by LFT, for feature space reduction is not particularly novel, we
The use of this trainable feature transformation layer enables
flexible experimentation as both ontology and molecular feature for
both drugs and proteins are reduced to the same dimensionality for
varying sizes of inputs; this allows for a high amount of modularity
across different experimental setups by adding different kinds of
features into the model. Additionally, the generated features may be
used for other tasks. % ; for
% example, the ontology LFT can be reused for a variety of DL2vec-based
% features with respect to other ontologies and hypotheses
We follow the results of \textit{DL2vec} \cite{} and use as
activation function $\sigma := \mathrm{LeakyReLU}$ which leads to
improved performance compared to other activation functions.

@@

We include these molecular and ontology-based sub-models within a
graph neural network (GNN) \cite{}. The graph underlying the GNN is
based on the protein--protein interaction (PPI) graph. The PPI dataset
is represented by a graph $G=(V,E)$, where each protein is represented
by a vertex $v\in V$, and each edge $e\in E\subseteq V\times V$
represents an interaction between two proteins. Additionally, we
introduce a mapping $x:V\rightarrow\mathbb{R}^{d}$ projecting each
vertex $v$ to its node feature $x_v := x(v)$, where $d$ denotes the
dimensionality of the node features.

@@

\subsubsection{Combined prediction model}
Combining half-twin and graph convolutional neural networks, we map
all protein representations to their respective node features,
initializing the graph convolutional update steps. The resulting
representations are used for a similarity prediction.
% as presented in
% figure ~\ref{fig:HalfTwinNetwork}
When combining ontology and molecular features with or without the
graph model, we concatenate both protein features and both drugs
features, before plugging them into the graph model for the similarity
computation. An overview of the model architecture, combining both
feature types, is shown in Supplementary Figure
\ref{fig:FullModelAllLayers}\todo{Move to supplement}. Here
the original representations are transformed by a dense layer and then
used as input of a stack (with height 3) of residual graph
convolutional blocks.

@@

In some cases, the $MicroAUC$ score may not be defined as in some
datasets some proteins or drugs have no interactions, leading to an
infeasible $TPR=0$ for all thresholds and an undefined $AUROC$ score
for that entity. For those entities, we impute the $MicroAUC$
interpolating linearly, by using the accuracy for this
subset:

\begin{equation*}
	\textrm{MicroAUC}_p(l,y) := 
	\begin{cases}
		MicroAUC'_p(l,y) & \text{if }\sum_{d_i\in
                  D}l(d_i,p)\neq 0\\
		Accuracy(l,y)&otherwise
	\end{cases}
\end{equation*}
Drugs and proteins can be interchanged in this formulation. We refer
to the different measures as protein-centric microAUC ($MicroAUC_p$)
and a drug-centric microAUC ($MicroAUC_d$).

@@

The results of our experiments are summarized in Table
\ref{tab:comparison1}; we calculated the performance of all compared
methods over their original splitting scheme and over a protein
split. We find that there is a large difference in performance when
evaluating over a drug--target pair split compared to a protein split,
with generally higher performance achieved when using the drug--target
pair split.  Second, when evaluating the same methods over a protein
split, we find a substantial performance difference in comparison to
the splitting scheme used in the original evaluation of each
method. DTI-CDF was originally evaluated on all three splitting
schemes underlining this point \cite{}. While \name{} provides
comparable performance to the na\"ive predictor and DTI-CDF in terms
of MacroAUC, it yields considerably better results with respect to
$\textrm{MicroAUC}_p$. We also find that methods that are trained
using a protein split generally result in higher $\textrm{MicroAUC}_p$
than methods trained using a drug--target pair split, indicating that
they may generalize better to unseen protein targets whereas methods
trained on a drug--target split potentially exploit hidden biases and
therefore generalize less well.

@@

As the difference in performance with different splitting schemes is
quite large, we further evaluated additional drug--target interaction
and drug--target affinity prediction methods that were trained and
evaluation on other datasets.  Following the results of MolTrans
\cite{MolTrans2020}, we reevaluated DeepDTI \cite{DeepDTI2017},
DeepDTA \cite{DeepDTA2018}, DeepConv-DTI \cite{DeepConvDTI2019}, and
MolTrans itself on the BioSnap dataset \cite{} and compared it to our
``na\"ive'' predictor as well as \name{} (see Table
\ref{tab:comparison2}).  MolTrans was evaluated over the drug--target
pair and the protein split; we were able to reproduce the MolTrans
results (Table \ref{tab:comparison2}), showing a substantial
difference based on the splitting scheme. We additionally computed the
$MicroAUC_p$ score for all considered methods, leading to similar 
results as observed on the Yaminishi dataset.

@@

There are other types of indirect features that could be added to our
model. A common feature that may be added are drug indications which are
predictive of drug--target interactions \cite{}; however, we do
not include them in our model as including drug indications would
allow our model to make many trivial predictions based only on
remembering which targets are often used for which indication;
including network information would likely benefit predictions based
on drug indications because different drugs may target the same
pathway through different mechanisms.

Combining bottom-up and top-down approaches in a single model can
follow different strategies. Interaction networks are used widely to
determine indirect effects of molecular changes \cite{} and predict
drug--target interactions \cite{}. Our work relies on graph neural
networks as a way to combine qualitative information about
interactions with additional features (molecular interaction,
phenotypic and functional features); even if only some of these
features benefit from the information the graph provides, graph neural
networks will allow further extension of our model with additional
features in the future.

@@

One major component of our experiments was to determine how the
information that is available to a machine learning model during
training affects the performance of the model. Similarly to previous
work \cite{}, we find significant differences in predictive
performance across different splitting schemes.

@@

% \subsection{Tested hypotheses}

% In this work we are testing the following hypotheses:
% \begin{enumerate}
% 	\item Can we build a model that outperforms state of the art approaches, combining top-down and bottom-up approaches?
% 	\item Are interaction networks sufficient to improve the performance of simple molecular predictors?
% \end{enumerate}
% We will test the first hypothesis by building a model that takes both top-down and bottom-up features into account. Thus, we propose a novel approach to combine those mutual exclusive attempts, through the usage of interaction networks, similarity and molecular features. Additionally, we test the latter by building a simple molecular DTI predictor and enhance it under usage of the interaction networks.\\

% \section{Methods}

% \subsection{Models}
% The used model consists of two separate models, that help to fuse together the two methods:
% \begin{enumerate}
% 	\item The molecular predictor
% 	\item The interaction network based predictor
% \end{enumerate}
% We build the molecular predictor by using pretrained, molecular fingerprints models for both drugs and proteins. Regarding proteins, we used the pretrained feature generator from \textit{DeepGoPlus} (\citep{DeepGoPlus}) that was originally designed for protein function prediction and is regarded as state of the art for this purpose. For drugs we used a pretrained fingerprint model from \textit{SMILES transformer} (\cite{SmilesTransformer}), that provides a simple and fast method to compute fingerprints through autoencoder models. The encodings from these two models were funneled into a simple deep neural network (see Figure~\ref{fig:01}) with few fully connected. \\
% The results of that prediction flow into the annotation of the protein-protein interaction (PPI) graph as depicted in (IMAGE). Hereby, the predictions of the molecular predictor are used as node features for the graph, with respect to the given drug. Thus, given a compound-target pair, the nodes of the PPI graph now hold bottom-up features, which can now be processed by the graph learning algorithms. \\
% The PPI graph is processed by different graph convolutional layers, that may underline the importance of either patterns or regions within the graph, to obtain a feature vector for the wanted node. In contrast to learning over whole graphs we perform node classification within the graph. These layers are either graph convolutional layers, that learn a certain kernel over the graph, or attention based. Different layers of both and other types such as were tested.  \\
% The drug-drug interaction features are retrieved by choosing the corresponding row in the adjacency matrix of the graph, thus leading to quite simple features. \\
% For the semantic similarity feature, that once again represents a top-down attribute, we artificially link each drug to its corresponding side effects in the MedDRA hierarchy. Concerning this hierarchy, drug-drug similarity is computed by the Resnik similarity (\cite{Resnik1995}). For the given compound we take the corresponding row of this symmetric similarity matrix. \\

% Thereby, we concatenate these three features together and funnel them into another deep neural network as depicted in figure \ref{fig:02}. This network finally yields our prediction. We hereby perform splits over both drugs and proteins, in order to test and show the discrepancy and increasing difficulty.\\ Implementation was done in PyTorch (\cite{Pytorch}) and is available on Github under \href{github.com/thinnerichs/KAUST-dti-metabol}{github.com/thinnerichs/KAUST-dti-metabol}. Graph learning methods were build with help of PyTorch-Geometric (\cite{PytorchGeometric}), a geometric deep learning extension library for PyTorch, that recently got a lot of attention in the machine learning community. This library gives the potential to use many state of the art graph learning mechanisms, such as plain but effective graph convolution (\citet{GCNConv}), Chebychev kernels (\cite{ChebConv}), ARMA kernels (\cite{ARMAConv}), translation-invariant operators (\cite{FeaStConv}), attention mechanisms (\cite{GATConv}), random walks (\cite{APPNPConv}) and mixtures of the latter two (\cite{SAGEConv}). The performance of these various layer types were tested for this particular problem, as discussed in the results section.

% \begin{figure}[!tpb]%figure1
% 	\centerline{\includegraphics[width=0.4\textwidth]{figures/MolecularPredictor.png}}
% 	\caption{Molecular predictor based on the generated features from DeepGoPlus and SMILES transformer.}\label{fig:MolPred}
% \end{figure}

% \begin{figure}[!tpb]%figure2
% \centerline{\includegraphics[width=0.4\textwidth]{figures/InteractionNetwork.png}}
% \caption{Deep neural network that predicts based on drug-drug interaction features and semantic similarity features over side effects for drugs, and graph convolution over protein-protein interaction networks for proteins. Protein and drug features are represented by blue and red, respectively.}\label{fig:02}
% \end{figure}

%\section{Discussion}

>>>>

\documentclass{bioinfo}
\usepackage[english]{babel}
\copyrightyear{2021} \pubyear{2021}
\usepackage{natbib}
\bibliographystyle{natbib}
\usepackage{url}

@@

Both approaches may be used in conjunction with network inference
\todo[inline]{CITE}\cite{}. Biological networks used for DTI prediction include
protein--protein interaction networks \todo[inline]{CITE}\cite{} and networks including
several other types of biological relations, including similarity
between represented entities \cite{Gottlieb2011}. Network-based DTI
prediction methods use the guilt-by-association principle
\cite{Oliver2000}
and assume that a protein is a likely target for a drug if many of the
protein's neighbors in the interaction network are targets of the drug
\cite{Gillis2012}. Network-based methods have been applied successfully
to DTI prediction. However, if DTIs are taken as direct physical
interactions between a drug and protein, it remains an unresolved
question whether the network-based guilt-by-association hypothesis is
true, or whether an interaction of a drug and protein dys-regulates
several of the protein's interaction partners, and therefore resulting
in downstream effects that are not direct interactions but only
downstream consequences of an interaction.

@@

As we want to learn from the similarity of drug side effects and
protein phenotypes, we use a deep half-twin neural network with a
contrastive loss using cosine similarity.  A half-twin neural network
aims to learn a similarity between two embeddings of the same
dimension.  The precomputed embeddings may have different
dimensionality and we first process them using a fully connected
neural network layer which takes as input an embedding and outputs a
representation of a particular size, i.e., we use this layer as a
trainable feature transformation and apply it to reduce the
representation size of the embeddings for drugs and proteins
separately.  An example structure for both types of features can be
found in Figure~\ref{fig:HalfTwinNetwork}.
% While a regular deep neural network, denoted by LFT, for feature space reduction is not particularly novel, we
The use of this trainable feature transformation layer enables
flexible experimentation as both ontology and molecular feature for
both drugs and proteins are reduced to the same dimensionality for
varying sizes of inputs; this allows for a high amount of modularity
across different experimental setups by adding different kinds of
features into the model. Additionally, the generated features may be
used for other tasks. % ; for
% example, the ontology LFT can be reused for a variety of DL2vec-based
% features with respect to other ontologies and hypotheses
We follow the results of \textit{DL2vec} \cite{DL2vec2020} and use as
activation function $\sigma := \mathrm{LeakyReLU}$ which leads to
improved performance compared to other activation functions.

@@

We include these molecular and ontology-based sub-models within a
graph neural network (GNN) \cite{GCNConv}. The graph underlying the GNN is
based on the protein--protein interaction (PPI) graph. The PPI dataset
is represented by a graph $G=(V,E)$, where each protein is represented
by a vertex $v\in V$, and each edge $e\in E\subseteq V\times V$
represents an interaction between two proteins. Additionally, we
introduce a mapping $x:V\rightarrow\mathbb{R}^{d}$ projecting each
vertex $v$ to its node feature $x_v := x(v)$, where $d$ denotes the
dimensionality of the node features.

@@

\subsubsection{Combined prediction model}
Combining half-twin and graph convolutional neural networks, we map
all protein representations to their respective node features,
initializing the graph convolutional update steps. The resulting
representations are used for a similarity prediction.
% as presented in
% figure ~\ref{fig:HalfTwinNetwork}
When combining ontology and molecular features with or without the
graph model, we concatenate both protein features and both drugs
features, before plugging them into the graph model for the similarity
computation. An overview of the model architecture, combining both
feature types, is shown in Supplementary Figure
\ref{fig:FullModelAllLayers}. Here
the original representations are transformed by a dense layer and then
used as input of a stack (with height 3) of residual graph
convolutional blocks.

@@

In some cases, the $MicroAUC$ score may not be defined as in some
datasets some proteins or drugs have no interactions, leading to an
infeasible $TPR=0$ for all thresholds and an undefined $AUROC$ score
for that entity. As this is quite common for DTI datasets, we do not omit but impute the $MicroAUC$
interpolating linearly for those entities, by using the accuracy for this
subset:

\begin{equation*}
	\textrm{MicroAUC}_p(l,y) := 
	\begin{cases}
		MicroAUC'_p(l,y) & \text{if }\sum_{d_i\in
                  D}l(d_i,p)\neq 0\\
		Accuracy(l,y)&otherwise
	\end{cases}
\end{equation*}
Note that drugs and proteins can be interchanged in this formulation. We refer
to the different measures as protein-centric microAUC ($MicroAUC_p$)
and a drug-centric microAUC ($MicroAUC_d$).

@@

The results of our experiments are summarized in Table
\ref{tab:comparison1}; we calculated the performance of all compared
methods over their original splitting scheme and over a protein
split. We find that there is a large difference in performance when
evaluating over a drug--target pair split compared to a protein split,
with generally higher performance achieved when using the drug--target
pair split.  Second, when evaluating the same methods over a protein
split, we find a substantial performance difference in comparison to
the splitting scheme used in the original evaluation of each
method. DTI-CDF was originally evaluated on all three splitting
schemes underlining this point \cite{DTI-CDF2019}. While \name{} provides
comparable performance to the na\"ive predictor and DTI-CDF in terms
of MacroAUC, it yields considerably better results with respect to
$\textrm{MicroAUC}_p$. We also find that methods that are trained
using a protein split generally result in higher $\textrm{MicroAUC}_p$
than methods trained using a drug--target pair split, indicating that
they may generalize better to unseen protein targets whereas methods
trained on a drug--target split potentially exploit hidden biases and
therefore generalize less well.

@@

As the difference in performance with different splitting schemes is
quite large, we further evaluated additional drug--target interaction
and drug--target affinity prediction methods that were trained and
evaluation on other datasets.  Following the results of MolTrans
\cite{MolTrans2020}, we reevaluated DeepDTI \cite{DeepDTI2017},
DeepDTA \cite{DeepDTA2018}, DeepConv-DTI \cite{DeepConvDTI2019}, and
MolTrans itself on the BioSnap dataset \cite{BioSnap2018} and compared it to our
``na\"ive'' predictor as well as \name{} (see Table
\ref{tab:comparison2}).  MolTrans was evaluated over the drug--target
pair and the protein split; we were able to reproduce the MolTrans
results (Table \ref{tab:comparison2}), showing a substantial
difference based on the splitting scheme. We additionally computed the
$MicroAUC_p$ score for all considered methods, leading to similar 
results as observed on the Yaminishi dataset.

@@

There are other types of indirect features that could be added to our
model. A common feature that may be added are drug indications which are
predictive of drug--target interactions \todo[inline]{CITE}\cite{}; however, we do
not include them in our model as including drug indications would
allow our model to make many trivial predictions based only on
remembering which targets are often used for which indication;
including network information would likely benefit predictions based
on drug indications because different drugs may target the same
pathway through different mechanisms.

Combining bottom-up and top-down approaches in a single model can
follow different strategies. Interaction networks are used widely to
determine indirect effects of molecular changes \todo[inline]{CITE}\cite{} and predict
drug--target interactions \todo[inline]{CITE}\cite{}. Our work relies on graph neural
networks as a way to combine qualitative information about
interactions with additional features (molecular interaction,
phenotypic and functional features); even if only some of these
features benefit from the information the graph provides, graph neural
networks will allow further extension of our model with additional
features in the future.

@@

One major component of our experiments was to determine how the
information that is available to a machine learning model during
training affects the performance of the model. Similarly to previous
work \cite{MolTrans2020}\cite{DeepConvDTI2019}, we find significant differences in predictive
performance across different splitting schemes.

@@
